---
title: "Privacy Preserving Homomorphic Linear Regression"
author: Kee Siong Ng (adapted by Balasubramanian Narasimhan)
date: '2018-01-01'
bibliography: ../bibtex/ipam2018.bib
link-citations: true
categories: []
tags: []
---

## Problem Setup

This example, based on the `homomorpheR` package, was communicated to
me by Kee Siong Ng. 

- `FIU` holds the response vector `y`
- `RE1` holds the data `x[, 1:3]`
- `RE2` RE2 holds the data `x[,4]`

Together, they want to learn a linear model to predict `y` using `x`
but in such a way that

- nobody sees each other's data,
- `RE1` holds the coefficients for the variables it holds, which is
  not visible to others
- `RE2` holds the coefficients for the variables it holds, which is
  not visible to others
- Prediction of a new instance requires the collaboration of `RE1` and `RE2`

`FIU` first generates a public-private key pair using the Paillier
scheme. The public key is used by `RE1` and `RE2` to encrypt their
data and do the computations. The private key is visible only to `FIU`
and is used by the `FIU` to decrypt data sent from the `RE`s.


## A Synthetic Data Set

```{r}
n <- 50
x1 = 11:(10+n)
x2 = runif(n,5,95)
x3 = rbinom(n,1,0.5)

x = data.frame(x1,x2,x3)
x = scale(x)
x = data.frame(1,x)    # the intercept is handled using a column of 1's

sigma = 1.4
eps = rnorm(x1, 0, sigma)  # generate noise vector

b = c(17, -2.5, 0.5,-5.2)   # the real model coefficient

y = b[1] + b[2] * x[,2] + b[3] * x[,3] + b[4] * x[,4] + scale(eps)  # target variable
```

## Batch Gradient Descent

This example uses a batch gradient descent algorithm for fitting the
model.

```{r}
lm_sgd <- function(iter, rate) {
  theta = c(0,0,0,0)
  alpha = rate
  for (iter in 1:iter) {
     adj = c(0,0,0,0)
     for (i in 1:4) {
         for (j in 1:n) {
             adj[i] = adj[i] + (sum(x[j,] * theta) - y[j]) * x[j,i]
         }
         # theta[i] = theta[i] - (alpha / n) * adj[i]         # adjust as we go is faster
     }
     for (i in 1:4) theta[i] = theta[i] - (alpha / n) * adj[i]
     print(adj)
  }
  print(theta)
}
```

As a check, we can run 100 iterations to see if it works.

```{r}
lm_sgd(100, 0.1)
```

## Some Tools

Some tools are needed for performing computations using real
numbers. Encryption schemes all work on large integers. 

```{r}
library(homomorpheR)

keypair = PaillierKeyPair$new(modulusBits = 1024)
pubkey = keypair$pubkey
privkey = keypair$getPrivateKey()

# Convenience function
zipf = function(x1,x2,x3) {
  res = c()
  for (i in 1:length(x1)) res = c(res,x1[i],x2[i],x3[i])
  res
}

# zipf(5,10,1)
# zipf(c(1,2,3), c(4,5,6), c(1,1,1))

######################################################################
# Paillier is defined only on positive integers.
# We approximate a floating point number X by 3 integers (a,b,c)
# such that X is approximately equal to (a + b/largenum) / 10^c
#
# Example: Assuming largenum = 2^10, 5.2 can be represented by
#  (5, 204, 0) = (5 + 204/1024) / 1 = 5.19922
#  or (50, 2040, 1) = (50 + 2040/1024) / 10 = 5.19922
#
# In general, the larger largenum is, the better the approximation.
#################

largenum = 2^50
encodeFloat = function(x, e=0) {
  x = x * 10^e
  x1 = floor(x)
  x2 = x - x1
  x2 = floor(x2 * largenum)
  exps = rep(e, length(x))
  zipf(x1,x2,exps)
}

# encodeFloat(5.2)
# encodeFloat(5.2, 2)
# encodeFloat(c(4.5,5.2,7), 1)

# Some useful constants
zero = pubkey$encrypt('0')
one = pubkey$encrypt('1')

############################################################################
# We encrypt each number (a,b,c) by encrypting a and b with the public key
# The input can be a vector of numbers
################

encrypt = function(z, e=0) {
  y = encodeFloat(z, e)
  res = rep(zero, length(y))
  for (i in 1:length(y)) {
      if (i %% 3 == 0) { res[i] = y[i] }
      else { res[i] = pubkey$encrypt(as.character(y[i])) }
  }
  res
}

############################################################################
# We decrypt each encrypted number (enc(a), enc(b),c) using the private key
# by computing (dec(enc(a)) + dec(enc(b))/largenum) / 10^c
# The input can be a vector of numbers.
# We need to deal with negative numbers too.
##################

decrypt = function(en) {
  uen = en
  for (i in 1:length(uen)) {
      if (i %% 3 != 0) uen[i] = privkey$decrypt(en[i])
  }
  # deal with negative results;
  # see https://crypto.stackexchange.com/questions/19457/how-can-i-do-minus-on-plaintexts-in-the-paillier-cryptosystem
  for (i in 1:length(uen)) {
    if (i %% 3 != 0 &&    # don't process exponents
        uen[i] >= as.double(pubkey$n/3.0)) { uen[i] = uen[i] - pubkey$n }
  }
  res = c()
  for (i in seq(1, length(uen), 3)) {
      res = c(res,
             (as.double(uen[i]) + as.double(uen[i+1]/(largenum))) / as.double(10^uen[i+2]))
  }
  return(res)
}

# decrypt(encrypt(4.7))
# decrypt(encrypt(0.5))
# decrypt(encrypt(-2.3))
# (z = encrypt(c(5.2,4.7,-2.6)))
# decrypt(z)

############################################################################
# We next define addition of two encrypted numbers: (enc(a1), enc(b1), c1) and (enc(a2), enc(b2), c2).
# Addition can be done on the encrypted a's and encrypted b's if c1 = c2. We make c1 = c2 true by
# multiplying by powers of 10 when necessary.
# The need to make c1 = c2 is the reason why we can't have c1 and c2 encrypted.
# The function works on vectors.
###############

addenc = function(x, y) {
  res = x
  for (i in seq(1, length(x), 3)) {
       if (x[i+2] != y[i+2]) {  # if different exponent, make the exponents equal by multiplying
          res[i+2] = max(x[i+2], y[i+2])
          xdiff = res[i+2] - x[i+2]
          ydiff = res[i+2] - y[i+2]
          if (xdiff > 0) {
              x[i] = pubkey$mult(x[i], 10^(xdiff))
              x[i+1] = pubkey$mult(x[i+1], 10^(xdiff))
          }
          if (ydiff > 0) {
              y[i] = pubkey$mult(y[i], 10^(ydiff))
              y[i+1] = pubkey$mult(y[i+1], 10^(ydiff))
          }
       }
       res[i] = pubkey$add(x[i], y[i])
       # print(privkey$decrypt(c(x[i], y[i], res[i])))
       res[i+1] = pubkey$add(x[i+1], y[i+1])
  }
  res
}

############################################################################
# We next define subtraction of two encrypted numbers: (enc(a1), enc(b1), c1) and (enc(a2), enc(b2), c2).
# We simply multiply the second encrypted number by -1 and then add it to the first encrypted number.
############

subenc = function(x, y) {
  for (i in seq(1, length(x), 3)) {
      # y[i:i+1] = pubkey$mult(y[i:i+1], -1)  # this doesn't work, why?
      y[i] = pubkey$mult(y[i], -1)
      y[i+1] = pubkey$mult(y[i+1], -1)
  }
  addenc(x, y)
}

#############################################################################################
# We next define the multiplication of an encrypted number en = (enc(a),enc(b),c) by a scalar y
# Again, Paillier only allows y to be an integer. To deal with floating point numbers, we use
# en * y = en * (y_int + y_frac)
#        = en * y_int + (enc(a) * floor(y_frac * 10^n), enc(b) * floor(y_frac * 10^n), n)
# where n is a parameter with larger n leading to better approximation. We use n = 3 below.
#######

smultenc = function(x,y) {
  yint = floor(y * 1000)
  res = x
  for (i in seq(1, length(x), 3)) {
      res[i] = pubkey$mult(x[i], yint)
      res[i+1] = pubkey$mult(x[i+1], yint)
      res[i+2] =  x[i+2] + 3
  }
  res
}

################
# Here are some test cases
####
# five = encrypt(5.4)
# ten = encrypt(10.2)
# five
# ten
# decrypt(addenc(five,ten))
# decrypt(subenc(ten,five))
# decrypt(subenc(five,ten))
# decrypt(addenc(c(five,ten), c(ten,ten)))
#
# five = encrypt(5.4,1)
# ten = encrypt(10.2,2)
# decrypt(addenc(five,ten))
# decrypt(subenc(ten,five))
# decrypt(addenc(c(five,ten), c(ten,ten)))
#
# decrypt(smultenc(five, 4))
# decrypt(smultenc(ten, 6))
# decrypt(smultenc(c(five,ten), 4))
# decrypt(smultenc(c(five,ten), -4))
```


Here are the functions to setup the parties in the computation.

```{r}
x1 = c()  # these 3 vectors belong to RE1
x2 = c()
x3 = c()

re1setup = function() {
  x1 <<- encrypt(x[,1])
  x2 <<- encrypt(x[,2])
  x3 <<- encrypt(x[,3])
}

x4 = c() # this belong to RE2

re2setup = function() {
  x4 <<- encrypt(x[,4])
}

mastersetup = function() {
  re1setup()
  re2setup()
}
```

The function `master_grad` below computes the gradient: the difference between predicted values (encrypted) by `RE`s and the true values

Note the gradient is unencrypted. We could encrypt it if necessary.

```{r}
master_grad <- function(x) {
  xd = decrypt(x)
  xd - y
  # subenc(xd, y)
}
```

Function `re1pred` computes the partial predictions based on `RE1` data only. The values are encrypted.

```{r}
re1pred <- function() {
   # return (theta1 * x[,1] + theta2 * x[,2] + theta3 * x[,3])
  return (addenc(smultenc(x1, theta1),
          addenc(smultenc(x2, theta2),
                 smultenc(x3, theta3)))
          )
}
```


`RE2` takes the partial prediction of `RE1` and then add its
contribution to the prediction. The final prediction values are
encrypted.

```{r}
re2pred <- function(z) {
   # return (z + theta4 * x[,4])
  return (addenc(z, smultenc(x4, theta4)))
}
```

We need a variable to keep track of progress.

```{r}
cadj = c(0,0,0,0) # a variable we want to print to keep track of progress
```

`RE1` model update function and implements the gradient descent update formula.
It takes the gradient (unencrypted) from the master and then adjust
its own theta. The whole computation is assumed to take place within
`RE1` and no encryption is required. 

```{r}
re1update <- function(z) {
   theta1 <<- theta1 - (alpha / n) * sum(z * x[,1])
   theta2 <<- theta2 - (alpha / n) * sum(z * x[,2])
   theta3 <<- theta3 - (alpha / n) * sum(z * x[,3])
   cadj[1] <<- sum(z * x[,1])
   cadj[2] <<- sum(z * x[,2])
   cadj[3] <<- sum(z * x[,3])
}
```

`RE2` model update function which implements the gradient descent
update formula. 

```{r}
re2update <- function(z) {
   theta4 <<- theta4 - (alpha / n) * sum(z * x[,4])
   cadj[4] <<- sum(z * x[,4])
}
```

After all that, we can now run the privacy-preserving linear
regression algorithm. The algorithm can be sensitive to the learning
rate.

```{r}
pp_lm_sgd <- function(iter, rate) {

  theta1 <<- 0
  theta2 <<- 0
  theta3 <<- 0
  theta4 <<- 0
  alpha <<- rate

  mastersetup()  # encrypt the data and set up communication

  for (i in 1:iter) {
      p1 = re1pred()     # partial prediction
      px = re2pred(p1)   # full prediction
      grad = master_grad(px)  # compute gradient based on difference between true values and predicted values
      re1update(grad)    # update models independently
      re2update(grad)
      print(cadj)
  }
  print(c(theta1,theta2,theta3,theta4))
}
```

## Results

First the standard result, not private.

```{r}
model0 = lm_sgd(100, 0.1)
```

Next the privacy preserving version.

```{r}
model1 = pp_lm_sgd(100, 0.1)
```











