---
title: Quick Intro to R
author: Balasubramanian Narasimhan
date: '2018-01-01'
bibliography: ../bibtex/ipam2018.bib
link-citations: true
categories: []
tags: []
---

## About the R language, briefly

If you are used to traditional computing languages, you will find R
different in many ways. The basic ideas behind R date back four
decades (to the S language of Chambers, Becker and Wilks] and have a
strong flavor of exploration: one can grapple with data, understand
its structure, visualize it, summarize it etc. Therefore, a common way
people use R is by typing a command and immediately see the
results. (Of course, scripts can also be written and fed to R for
batch execution.)

The core of R itself is reasonably small, but over time, it has also
become a vehicle for researchers to disseminate new tools and
methodologies via packages. That is one reason for R's popularity:
there are thousands of packages (12000+ as of this writing) that extend
R in many useful ways.

The [CRAN](https://cloud.r-project.org) website is a crucial resource,
hosting all the software, documentation and manuals.

## Some detail on the language

A very brief run-through. 

### R Style and Nomenclature

- R variable names frequently contain periods and
underscores. Example: `male.cholesterol` or `male_cholesterol`

- R users tend to use the word _objects_ to refer to R variables,
functions, datasets, etc. This refers to something more than object in
an object-oriented programming style.

- John Chambers: Everything that exists in R is an object. Everything that happens is a function call.

So, in R, all action occurs via _functions_. Even something as
simple as

```{r}
1 + 2
```
is computed via a function call.

- You can use `=` or `<-` for assignment, but `<-` is better

### Vectors and Indexing

- _R uses 1-based indexing!_

- R has no concept of a scalar. A scalar is simply a vector of length
  1.

- R can handle both numeric and non-numeric (non-numeric)
   data. Beware R shows its statistical origins in that some
   non-numeric data may be automatically converted to _factors_. 

- Combining numeric and non-numeric into vectors causes silent
  coercion into the type that can accommodate the result!

```{r}
c(1, 2, "foo", "bar")
```

- Logicals in R are `TRUE` and `FALSE`.

- Indexing can be done with integers, names, logical values etc.
```{r}
x <- 1:10
x[x %% 2 == 0]
x[-(1:5)]
```

- Lists are versatile data structures that can grow or shrink and
contain heterogeneous data. They are constructed using the `list`
function:

```{r}
(aList <- list(a = 1, b = 2, c = list(1, 2, "abc")))
```
Note how a list prints differently. 

- With lists, the individual _elements_ can also be accessed using the
_dollar_ (`$`) notation.
```{r}
aList$c
```

- R has a notion of a missing value, `NA`. This is not the same as
  `NULL`, which indicates _nothing is present_. 

```{r}
miss1 <- c(1.0, NA, 2.0)
2 * miss1
```
Some numerical computations will have to provide hints on how to handle the
missing values. For example, the `mean` function computes the average
of a set of numbers.

```{r}
## No hint to process missing values
mean(miss1)
## Remove missing values before processing
mean(miss1, na.rm = TRUE)
```
One can check for missing-ness or nullity using the `is` family of
functions.

```{r}
is.null(c())
is.null(NA)
## This should produce a warning
is.na(c())
is.na(NA)
```
There are many others: `is.numeric`, `is.list`, `is.vector`, etc.

### Numerics

- Standard operations. When you perform arithmetic on vectors, the
operations happen on all elements.

```{r}
## Add two vectors
1:3 + 2:4
## Multiply a vector by 2
2 * 1:3
## Better to have parenthesis
2 * (1:3)
## Divide
c(2, 4, 6) / c(2, 4, 6)
## Halve
c(2, 4, 6) / 2
## R recycles shorter vector to match length
c(2, 4, 6, 8) / c(1, 2)
```
The last operation shows how R tries to make two vectors conform in
length and provides no warning. It is easy to avoid such constructs.

- The usual comparison operators are available: `==` for equality, `!=`
for not equal to, `>=` for greater than or equal to, etc.

```{r}
xx <- 1:3
xx == xx
## 1 is expanded to match length of xx
xx > 1
```

### Matrices

The function `matrix` can be used for creating matrices which are
two-dimensional arrays.

```{r}
## Create a 3 by 2 matrix.
m <- matrix(1:6, nrow = 3)
m
```

Another way is to use existing vectors to _bind_ into a matrix.
```{r}
xx <- 1:3
yy <- 4:6
## Bind by columns
(m2 <- cbind(xx, yy))
## Bind by rows
(m3 <- rbind(xx, yy))
```

The matrix `m2` has the same content as `m` above, but the columns
have names `xx` and `yy` which can be used in subsetting/indexing.

```{r}
## Access element in row 1, column 2
m[1, 2]
## Access second column
m[ , 2]
## Do the same with matrix m2
m2[, "yy"]
## Access the third row of m
m[3, ]
```

- Sparse Matrices are provided by specialized packages such as
  `Matrix` or `slam`. 

- R provides a large suite of functions for numerical analysis,
  optimization etc. 
  
## R Packages

[CRAN](https://cloud.r-project.org) hosts over 12K packages. Not all
are of great quality.  

- CRAN Taskviews provide an excellent summary of packages useful for a
  specific area.
  
- We will focus on two packages
  [`opencpu`](https://cran.r-project.org/package=opencpu) and 
  [`CVXR`](https://cran.r-project.org/package=CVXR) that we will use
  for demonstrations.
  
  
