---
title: Quick Intro to R
author: Balasubramanian Narasimhan
date: '2018-01-01'
bibliography: ../bibtex/ipam2018.bib
link-citations: true
categories: []
tags: []
---

## About the R language, briefly

If you are used to traditional computing languages, you will find R
different in many ways. The basic ideas behind R date back four
decades and have a strong flavor of exploration: one can grapple with
data, understand its structure, visualize it, summarize it
etc. Therefore, a common way people use R is by typing a command and
immediately see the results. (Of course, scripts can also be written
and fed to R for batch execution.)

The core of R itself is reasonably small, but over time, it has also
become a vehicle for researchers to disseminate new tools and
methodologies via packages. That is one reason for R's popularity:
there are thousands of packages (12000+ as of this writing) that extend
R in many useful ways.

The [CRAN](https://cloud.r-project.org) website is something you
will consult frequently for both the software, documentation and
packages others have developed.

## Some detail on the language

Consider the following code snippet which creates a few variables.

```{r}
x <- 1
y = 2
z <- x + y
```

Like other computer languages, R has ways of naming things in the
language. Above, we used `x` as a name for the value 1 and `y` for the
value 2. The names have to follow some rules. It is sufficient to be
aware that they must start with an alphabetic character and can
contain periods and underscores. Also, for obvious reasons, space is
not permitted. (It is common to see names for variables such as
`male.cholesterol` or `male_cholesterol`!)

__Nomenclature__: R users tend to use the word _objects_ to refer to R
variables, functions, datasets, etc.


John Chambers: 

- Everything that exists is an object.
- Everything that happens is a function call.

So, in R, all action occurs via _functions_. Even something as
simple as

```{r}
1 + 2
```
is computed via functions.

### Should I use `=` or `<-` for assignment?

In R, both `=` and `<-` can be used for assigning a value to
variables. However, using `<-` (which I do) is the best thing to do.


### Vectors and Indexing

1. R has no concept of a scalar. A scalar is simply a vector of length 1.

2. R can handle both numeric and non-numeric (non-numeric)
data.

```{r}
x <- c(1.0, 2)
x
typeof(x)
```


```{r}
y <- c("abc", "d", "e", 'fgh')
y
typeof(y)
```

```{r}
z <- 1:5
z
typeof(z)
```

```{r}
w <- c(TRUE, FALSE, TRUE, TRUE)
w
typeof(w)
```

The `c` stands for the _combine_ function and it creates a vector of
two numbers for `x` and a vector of four strings for `y`. Note how
both single and double quotes may be used (useful when we have quotes
within strings). For `z` we use a shortcut `1:5` for creating a
sequence of integers from 1 to 5. And finally, `w` is a logical
vector; R recognizes the symbols `TRUE` and `FALSE` as special
symbols; you cannot have a variable named `TRUE` for example! (The
`typeof` function is useful to understand basic underlying types.)

3. _R uses 1-based indexing!_

Often, one needs to access a part, or a subset or a slice of a
vector. This is done by specifying indices indexing construct

```{r}
## The first element; indexing begins from 1
x[1]
## The third element of y
y[3]
## The second to fourth element of z
z[2:4]
## The first and last element of y
y[c(1, length(y))]
```

Note the use of the function `length` that returns the length of `y`
(4 for us).

4. Nothing stops one from combining types.

```{r}
## Combine x and y into one
c(x, y)
```

Note, however, that the last combine operation silently coerces
everything to strings. This is because vectors contain _homogeneous_
elements. That seems limiting, because sometimes you may have both
types of data and you don't want to be converting things back and
forth.

5. Lists are versatile data structures that can grow or shrink and
contain heterogeneous data. They are constructed using the `list`
function:

```{r}
aList <- list(1, 2, list(1, 2, "abc"))
aList
```
Note how a list prints differently. Individual elements of the list,
unlike the vectors above, are accessed using the double bracket
notation, suggested by the printing. Note also that there is no
coercion of types.

```{r}
## The second element
aList[[2]]
## The third element, which is itself a list!
aList[[3]]
## The second element of the third element
aList[[3]][[2]]
```

With lists, the single bracket indexing behaves differently from
double bracket indexing.

```{r}
aList[[2]]
aList[2]
```
The difference is clear from the way each is printed: the former is
just the second element of the list whereas the latter is another
list whose second element is from the original list.

The rule is simple: single bracket indexing returns the same type of
object.

```{r}
typeof(aList[[2]])
typeof(aList[2])
```

6. Negative indexing is a convenient way to drop some elements from a
vector.

```{r}
## Drop the first element of x
x[-1]
## Drop the last element of y
y[-length(y)]
## Drop the first and last element of aList
aList[c(-1, -length(aList))]
```

Mixing of negative and non-negative indices is not permitted.

```{r, error = TRUE}
## This results in an error
y[c(-1, 3:4)]
```

R also allows logical indexing:

```{r}
## Select y elements where w is TRUE
y[w]
```
will select the first, third and fourth elements and drop the
rest. Selecting elements based on conditions is very useful and we
will see further examples.

8. Missing and `NULL` values

R has a notion of a missing value that can be used to indicate data is
missing for some cases, an all too real phenomenon. It is denoted by
`NA`.

```{r}
miss1 <- c(1.0, NA, 2.0)
2 * miss1
```
Notice how the last operation did the appropriate thing with the
missing value. It is extremely convenient to be able to use missing
values as you would any other object in R. But numerical computations
will have to provide hints on how to handle the missing values. For
example, the `mean` function computes the average of a set of numbers.

```{r}
## No hint to process missing values
mean(miss1)
## Remove missing values before processing
mean(miss1, na.rm = TRUE)
```

Another value `NULL` is used to indicate _nothing is present_. Note
that it is semantically different from a missing value.

```{r}
NULL
## Combine nothing
c()
```
One can check for missing-ness or nullity using the `is` family of
functions.

```{r}
is.null(c())
is.null(NA)
## This should produce a warning
is.na(c())
is.na(NA)
```
There are many others: `is.numeric`, `is.list`, `is.vector`, etc.

9. Arithmetic and logical operations

The standard operations are all available: `+`, `-`, `*`
(multiplication), `/` division. In R, when you perform arithmetic on
vectors, the operations happen on all elements.

```{r}
## Add two vectors
1:3 + 2:4
## Multiply a vector by 2
2 * 1:3
## Better to have parenthesis
2 * (1:3)
## Divide
c(2, 4, 6) / c(2, 4, 6)
## Halve
c(2, 4, 6) / 2
## R recycles shorter vector to match length
c(2, 4, 6, 8) / c(1, 2)
## Above is same as
c(2, 4, 6, 8) / c(1, 2, 1, 2)
## Warning, but not error below
c(2, 4, 6) / c(1, 2)
```
The last operation shows how R tries to make two vectors conform in
length and provides a warning. _Good code avoids relying on such
behaviors as they can cause unpredictable errors._

The usual comparison operators are available: `==` for equality, `!=`
for not equal to, `>=` for greater than or equal to, etc.

```{r}
xx <- 1:3
xx == xx
## 1 is expanded to match length of xx
xx > 1
```

10. Comparison operators can be used to select subsets of vectors.  Some
examples with the understanding that `a %% 2` returns the reminder
upon division of a by 2.

```{r}
xx <- 1:10
## Pick all numbers >= 5
xx[ xx >= 5]
## Pick even numbers from 1 to 10
xx[ xx %% 2 == 0]
## Pick odd numbers from 1 to 10
xx[ xx %% 2 != 0]
```

11. Coercion

We saw above that some functions, can silently coerce the results to
something meaningful. In many case, such coercions can be useful.

How many even numbers between 1 and 10?

```{r}
xx <- 1:10
xx %% 2 == 0
sum(xx %% 2 == 0)
```
Here `xx %% 2 == 0` is a list of 10 logical values with `TRUE`
wherever we have an even number. The function `sum` converts `TRUE`
values to 1 and `FALSE` values to 0 and results to provide the answer.

R usually coerces the results where possible to the type that can
accomodate the result. If it cannot, it signals an error.

There are many explicit coercion functions such as `as.numeric`,
`as.integer`, `as.list`.

```{r}
xx <- 1:5
as.integer(xx %% 2 == 0)
as.character(xx)
as.list(xx)
```

Coercion functions are useful when processing external data for
computational work.


12. Naming

R allows one to add _names_ to objects.

```{r}
named_x <- c(a = 1.02, b = 2, 3)
named_x
```
Above, only two of the three elements were named. This makes the third
element have an empty name. The function `names` allows one to
retrieve the names of an object.

```{r}
names(named_x)
```
The naming facility allows one to access elements of vectors using
names rather than indices.

```{r}
## Equivalent to named_x[2]
named_x["b"]
## Equivalent to named_x[1:2]
named_x[c("a", "b")]
```

_Naming is an extremely useful tool in writing readable code_. One
might worry about a performance penalty but it is negligible in most
cases and the gains in readability far outweigh any inefficiencies.

Naming works for lists too.

```{r}
named_list <- list(x = x, y = y, zed = z)
named_list[c("x", "zed")]
```

With lists, the individual _elements_ can also be accessed using the
_dollar_ (`$`) notation.

```{r}
named_list$zed
```

Much of R code and functions exploit naming; many functions return
more than one value and they are often stuffed into a named vector or
list.

```{r}
aSummary <- summary(1:10)
aSummary
names(aSummary)
typeof(aSummary)
aSummary["Median"]
```
