---
title: Vertically Partitioned Data
author: Balasubramanian Narasimhan
date: '2018-01-01'
bibliography: ../bibtex/ipam2018.bib
link-citations: true
categories: []
tags: []
---

## Introduction

@doi:10.1093/jamia/ocv146 implement a logistic regression using
 vertically partitioned data. 
 

## Logistic Regression

We first generate some synthetic data where we have 1000 observations
and 20 parameters.


```{r}
suppressMessages(suppressWarnings(library(CVXR)))
p <- 10
n <- 100
##p <- 5
##n <- 10
offset <- 0
sigma <- 3
DENSITY <- 0.2

set.seed(18321)
beta_true <- stats::rnorm(p)
X <- matrix(stats::rnorm(n*p, 0, 5), nrow = n, ncol = p)
y <- sign(X %*% beta_true + offset + stats::rnorm(n, 0, sigma))
Y <- (y+1)/2  ## on 0, 1 scale.
```

A basic logistic regression model can be fit using `glm`.

```{r, eval = FALSE}
g <- glm(Y ~ 0 + X, family="binomial")
y.pred <- ifelse(predict(g, newdata = as.data.frame(X), type="response") > 0.5, 1, 0)
knitr::kable(table(y, y.pred))
```

```{r, eval = FALSE}
suppressWarnings(suppressMessages(library(CVXR)))
beta <- Variable(p)
obj <- -sum(X[y <= 0, ] %*% beta) - sum(logistic(-X %*% beta))
prob <- Problem(Maximize(obj))
result <- solve(prob)
beta_value <- result$getValue(beta)
```

## Ridged Version

However, it is common to fit penalized models, such as ridged version
where a parameter $\lambda$ controls the smoothness.

Solve for $\lambda = 5.$

```{r}
beta <- Variable(p)
lambda <- 5
loss <- sum_entries(logistic(-mul_elemwise(y, X %*% beta))) + lambda/2 * sum_squares(beta)
problem <- Problem(Minimize(loss))
res <- solve(problem)
beta_primal <- res$getValue(beta)
```

## Vertical Partitioning

First check that the dual solves the same problem.  Here is the dual

```{r}
n <- length(y)
alpha <- Variable(n)
XX <- as.numeric(y) * X
tXX <- t(XX)
K <- XX %*% tXX
obj <- 0.5 * quad_form(alpha, K) / lambda - sum(entr(alpha) + entr(1-alpha))
vert_primal <- Problem(Minimize(obj))
result <- solve(vert_primal)
alphaValue <- result$getValue(alpha)
##print(alphaValue)
beta_dual <- tXX %*% alphaValue / lambda
##y * (X %*% beta)
cbind(beta_primal, beta_dual)
```

They should match.


## Split version.

Here the data is vertically split.


```{r}
splits <- list(1:3, 4:6, 7:10)
Xs <- lapply(splits, function(ind) X[, ind])
n <- length(y)
alpha <- Variable(n)
## Form XX and its transpose
XXs <- lapply(Xs, function(x) {
    XX <- as.numeric(y) * x
    tXX <- t(XX)
    list(XX, tXX)
})

K <- Reduce("+", lapply(XXs, function(l) l[[1L]] %*% l[[2L]]))

obj <- 0.5 * quad_form(alpha, K) / lambda - sum(entr(alpha) + entr(1-alpha))
vert_dual <- Problem(Minimize(obj))
result <- solve(vert_dual)
alphaValue <- result$getValue(alpha)
##print(alphaValue)

tXX <- do.call(rbind, lapply(XXs, function(x) x[[2]]))

##beta_dual_split <- rbind(tXX1, tXX2) %*% alphaValue / lambda
beta_dual_split <- tXX %*% alphaValue / lambda
##y * (X %*% beta)
cbind(beta_dual, beta_dual_split)
```

That also matches.


## References
